# Data Importing/Exporting and interaction with other programmes {#data-import}

This chapter is for code examples of data importing/exporting and interactions with other programmes and databases.

## Libraries

```{r message=FALSE}
library(tidyverse)
library(DBI)
library(dbplyr)
library(haven)
library(bigrquery)
```

## Navigating folders

A couple of pointers to navigate from your working directory, which, if you're using R projects (it is highly recommended that you do) will be wherever the `.Rproj` file is located 

### Down

To navigate down folders use `/`. The path given below saves the file **my_df.csv** in the **data** folder, which itself is inside the **monthly_work** folder
```{r eval=FALSE}
readr::write_csv(
  x = my_dataframe
  , path = "monthly_work/data/my_df.csv"
)
```

### Up

To go up a folder use `../`. In particular you may need to do this when running Rmarkdown files. Rmarkdown files use their location as the working directory. If you have created an **R** folder, say, to stash all your scripts in, and a **data** folder to stash your data files in, then you will need to go up, before going down...

The path below goes up one folder, then into the **data** folder, where the **lookup_table.csv** is located.
```{r eval=FALSE}
lookup_table <- readr::read_csv(
  file = "../data/lookup_table.csv"
)
```

## Working with files in R

This section focusses on reading in various filetypes for working on in R memory. 

### .rds

.rds is R's native file format, any object you create in R can be saved as a .rds file. The functions `readRDS` and `saveRDS` are base R functions.

```{r eval=FALSE}
saveRDS(
  object = my_model #specify the R object you want to save
  , file = "2019_0418_my_model.rds" #give it a name, don't forget the file extension
)
```


### .csv

We use the functions `read_csv` and `write_csv` from the `readr` package (which is part of the `tidyverse`). These are a little bit *cleverer* than their base counterparts, however, this cleverness can catch you out.

The file **messy_pokemon_data.csv** contains pokemon go captures data which has been deliberately messed up a bit. `read_csv` imputes the column specification from the first 1000 rows, which is fine if your first 1000 rows are representative of the data type. If not then subsequent data that can't be coerced into the imputed data type will be replaced with NA. 

Looking at the column specification below notice that `read_csv` has recognised **time_first_capture** as a time type, but not **date_first_capture** as date type. Given the information that **combat_power** should be numeric we can see that something is also amiss here as `read_csv` has guessed character type for this column.  
```{r}
pokemon <- readr::read_csv(
  file = "data/messy_pokemon_data.csv"
)
```

Let's have a quick look at some data from these columns
```{r}
pokemon %>% 
  dplyr::select(species, combat_power, date_first_capture, time_first_capture) %>% 
  dplyr::arrange(desc(combat_power)) %>% 
  head()
```

The pokemon dataset has less than 1000 rows so `read_csv` has 'seen' the letters mixed in with some of the numbers in the **combat_power** column. It has guessed at character type because everything in the column can be coerced to character type.

What if there are more than 1000 rows? For example, say you have a numeric column, but there are some letters prefixed to the numbers in some of the post-row-1000 rows. These values are still meaningful to you, and with some data wrangling you can extract the actual numbers. Unfortunately `read_csv` has guessed at type double based on the first 1000 rows and since character type cannot be coerced into double, these values will be replaced with `NA`. If you have messy data like this the best thing to do is to force `read_csv` to read in as character type to preserve all values as they appear, you can then sort out the mess yourself.

You can specify the column data type using the `col_types` argument. Below I have used a compact string of abbreviations to specify the column types, see the help at `?read_csv` or the `readr` vignette for the full list. You can see I got many parsing failures, which I can access with `problems()`. This is a data frame of the values that `read_csv` was unable to coerce into the type I specified, and so has replaced with NA. 
```{r}
pokemon <- readr::read_csv(
  file = "data/messy_pokemon_data.csv"
  , col_types = "cdddcdcccDt"
)
# c = character, d = double, D = Date, t = time
```
Let's take a look at the problems.
```{r}
problems(pokemon) %>% 
  head()
```

And since I know that there are problems with **combat_power** let's take a look there.
```{r}
problems(pokemon) %>% 
  dplyr::filter(col == "combat_power") %>% 
  head()
  
```
The `problems()` feature in `read_csv` is super useful, it helps you isolate the problem data so you can fix it.

Other arguments within `read_csv` that I will just mention, with their default settings are

- `col_names = TRUE`: the first row on the imput is used as the column names.
- `na = c("", "NA")`: the default values to interpret as `NA`.
- `trim_ws = TRUE`: by default trims leading/trailing white space. 
- `skip = 0`: number of lines to skip before reading data.
- `guess_max = min(1000, n_max)`: maximum number of records to use for guessing column type. NB the bigger this is the longer it will take to read in the data.

 


### .xls and .xlsx

Excel workbooks come in many shapes and sizes. You may have one or many worksheets in one or many workbooks, there may only be certain cells that you are interested in. Below are a few examples of how to cope with these variations using `readxl` and `purrr` to iterate over either worksheets and/or workbooks.

#### A single worksheet in a single workbook

The simplest combination, you are interested in one rectangular datset in a particular worksheet in one workbook. Leaving the defaults works nicely on this dataset, though perhaps "units" needs some investigating.

```{r}

data_from_xl <- readxl::read_excel(
  path = "data/port0499.xlsx"
)

tibble::glimpse(data_from_xl)

```

Let's set a few of the other arguments, run `?read_excel` in the console to see the full list.
```{r}
data_from_xl <- readxl::read_excel(
  path = "data/port0499.xlsx"
  , sheet = 1 #number or name of sheet, default is first sheet
  , col_names = TRUE #default
  , col_types = "text" #a single type will recycle to all columns, specify each using character vector of the same length eg c("numeric", "text", ...)
)

tibble::glimpse(data_from_xl)
```


#### A single worksheet in many workbooks

For example, you collect sea passenger count data from many ferry companies, the data all has the same struture and you want to read it in and row bind into a single dataframe in R. 

```{r}

```

#### Many worksheets in a single workbook

You have a single workbook, but it contains many worksheets of interest, each containing rectangular data with the same structure.
```{r}

```

#### Many worksheets in many workbooks

```{r}

```

#### Non-rectangular data in many workbooks

You have received some kind of data entry form that has been done in excel in a more human readable, rather than machine readable, format. Some of the cells contain instructions and admin data so you only want the data held in specific cells. This is non-rectangular data, that is, the data of interest is dotted all over the place.

```{r}

```


 
 - read .xls/s
 - read single tab
 - read many tabs from one workbook (FASTS)
 - read individual cells/cell ranges (SP)
 - write with formatting openxlsx/xltabr
 
 
### .sav

## Connecting to databases

### SQL
 - DBI
 - default schema
  - link to C&C talks on the subject

### GCP
 - BigQuery - bigrquery