```{r include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

# Working with dates {#dates}




```{r include=FALSE}
library(dplyr)
library(readr)
library(reshape2)
library(tidyr)
library(janitor)
library(lubridate)

# Read in required data using public data.gov extract
road_accidents <- readr::read_rds("data/road_accidents_2017.RDS")

# create character version of date to use in examples
road_accidents$Date1 <- as.character(road_accidents$Date)

```


This chapter provides an overview of working with dates, for example extracting year or month from a date, and converting characters to a date.

One of the main packages used to work with dates is **lubridate**.

More information can be found on the lubridates cheatsheet at the following link: https://www.rstudio.com/resources/cheatsheets/

Date vectors are just double vectors with an additional class attribute set as
"Date".  

**Note - The following borrows directly from Hadley's Advanced R site
so we need to attribute appropriately and use correct licence**

```{r}
today <- Sys.Date()
typeof(today)
attributes(today)
```

The value of the double (which can be seen by stripping the class), represents
the number of days since "1970-01-01"^[This special date is known as the Unix 
Epoch]:

```{r}
date <- as.Date("1970-02-01")
unclass(date)
```



## Converting a character to a date 

In R, dates can be converted to a specific date variable type in order to use the variable as a date.

Having a variable as a date means that you can:
 - extract the different elements of the date (year, month etc.)
 - calculate differences between dates

This can be done in the following way:

- Identify the order of the year, month, day and use the appropriate function (ymd, mdy, dmy etc.)

```{r, echo = TRUE}

# convert date to date object

# check class of date
class(road_accidents$Date1)

# look at the date variable and see what order it is in (year-m-d)
# therefore use the ymd function
road_accidents$Date1 <- lubridate::ymd(road_accidents$Date1)

# now check class
class(road_accidents$Date1)

```


## Get year from date

Use the **year** function from lubridate:

```{r}

road_accidents$Year <- lubridate::year(road_accidents$Date1)

```

## Get month from date

Use the **month** function from lubridate:

```{r}

road_accidents$Month <- lubridate::month(road_accidents$Date1)

```


## Get day from date

Use the **day** function from lubridate:

```{r}

road_accidents$Day <- lubridate::day(road_accidents$Date1)

```

## Get weekday from date

Use the **wday** function from lubridate to get the weekday label:

```{r}

road_accidents$weekday <- lubridate::wday(road_accidents$Date1)

```


## Get quarter from date

Use the **quarter** function from lubridate:

```{r}

road_accidents$Quarter <- lubridate::quarter(road_accidents$Date1)

```

```{r, include = FALSE}

dates <- dplyr::select(road_accidents, Date1, Year, Quarter, Month, Day, weekday)

```


```{r, echo = FALSE, results='asis'}

knitr::kable(head(dates),
caption = "Using lubridate to extract time information")

```


## Find difference between two dates


```{r, include = FALSE}

# First create new date column so difference between two dates can be found

road_accidents$Date2 <-round_date(road_accidents$Date1, unit = "month")

```

Use the **as.duration** function to find the duration between two dates. The duration to be measured can be specified:

- dhours
- dweeks
- ddays
- dminutes
- dyears

To find out the number of days difference, the **as.duration** function calculates the duration in seconds so the duration must be divided by the desired duration (ddays) to convert to duration in days.

```{r}

road_accidents$date_diff <- lubridate::as.duration(road_accidents$Date2 %--% road_accidents$Date1) / ddays(1)

```

The number after **ddays** indicates by how many units the duration is (i.e. one day).


## Convert month (interger to character)

Base R has a useful function which takes the month numbers and converts them to the corresponding text.

```{r}

road_accidents$Month_lab <- month.abb[road_accidents$Month]

```

## Convert month (character to interger)

Base R has a useful function which takes the month text and converts them to the corresponding number.

```{r}
 road_accidents$Month <- match(road_accidents$Month_lab,month.abb)

```


## Merge separate date information into a date

The lubridate package can be used in conjunction with the paste function to combine columns separate date information (e.g. year, month, day) into one date variable.

```{r}

road_accidents$date <- paste(road_accidents$Year, road_accidents$Month, road_accidents$Day, sep="-") %>% ymd() %>% as.Date()

```


## Working with date-times

A date-time stores date and time information.


## Converting a character to a date-time

This is similar to converting a character to a date as mentioned above.

This can be done in the following way:

- Identify the order of the year, month, day, and time elements (hour, minute and second or just hour and minute) and use the appropriate function (ymd, mdy, dmy etc.)

```{r, eval = false}

# convert date to date object

# look at the date variable and see what order it is in (year-m-d, hms "2017-11-28 14:00)
# therefore use the ymd_hm
road_accidents$Date_time1 <- lubridate::ymd_hm(road_accidents$Date_time)

```


## Extract date from date time variable

Use the date function to extract the date from a date time variable.

The year/month/day information can then be extracted from the date using the code examples above.

```{r, eval = false}


road_accidents$Date2 <- lubridate::date(road_accidents$Date_time)

```

