# Table/Data Frame manipulation {#tables}

```{r include=FALSE}
library(dplyr)
library(readr)
library(reshape2)
library(tidyr)

# Read in required data using public data.gov extract
road_accidents <- readr::read_rds("data/road_accidents_2017.RDS")

```


This chapter provides an overview of code examples for table or data frame manipulation.

One of the main things you will have to do in any R project or RAP project will be manipulating the data that you are using in order to get it into the format you require.

One of the main packages used to manipulate data is the dplyr package which we recommend and use throughout this book. 

Dplyr is built to work with **tidy data**. To find out more about tidy data please look at the following link https://r4ds.had.co.nz/tidy-data.html but the general principles are:

1. Each variables must have its own column
2. Each observation must have its own row
3. Each value must have its own cell

## Pivot and reshape tables

There will be two examples for pivoting tables provided:

* The tidyr package uses the gather/spread functions and is often used to create tidy data
* The reshape2 package is also a useful package to pivot tables and has added functionality such as providing totals of columns etc.


We want to have the day of the week variable running along the top so each day of the week is its own table.

```{r, echo = FALSE, results='asis'}
# Create smaller dataset for example
road_accidents_small <- road_accidents %>%
  dplyr::group_by(Accident_Severity, Day_of_Week) %>%
  dplyr::tally()

knitr::kable(head(road_accidents_small),
caption = "Number of road accidents by accident severity and weekday")

```



**Tidyr package**

Using the tidyr package, gather and spread functions can be used  to pivot the table views:

- **gather** makes wide data longer i.e. variables running along the top can be "gathered" into rows running down.

- **spread** makes long data wider i.e. one variable can be spread and run along the top with each value being a variable.

```{r}
# Pivot table using tidyr package
library(tidyr)
road_accidents_weekdays <- road_accidents_small %>%
  tidyr::spread(Day_of_Week, n)
  
```

With the spread function above you need to first specify the variable you want to spread, in this case Day_of_Week, and then the variable that will be used to populate the columns (n).

```{r, echo = FALSE, results='asis'}

knitr::kable(head(road_accidents_weekdays),
caption = "Number of road accidents by accident severity and weekday, tidyr::spread")

```


The opposite can also be done using the gather function: GET HELP ON THIS

```{r}
# Pivot table using tidyr package
library(tidyr)
road_accidents_gather <- road_accidents_weekdays %>%
  tidyr::gather(Accident_Severity, n, 1:7)
  
```


**Reshape2 package**

Again, this has two functions which can be used to pivot tables:

- **melt** makes wide data longer 

- **dcast** makes long data wider


```{r}
# Pivot table using reshape2 package
library(reshape2)
road_accidents_weekdays2 <- reshape2::dcast(road_accidents_small, Accident_Severity ~ Day_of_Week,
                                            value.var = "n")
  
```

With the dcast function above you need to first specify the variable(s) you want in long format (multiple variables seperated by "+"), in this case Accident_Severity, and then the wide format variable(s) are put after the tilda (again multiple seperated by "+"). The value.var argument specifices which column will be used to populate the new columns, in this case it is n.

```{r, echo = FALSE, results='asis'}

knitr::kable(head(road_accidents_weekdays2),
caption = "Number of road accidents by accident severity and weekday, reshape2::dcast")

```

If you want to create sums and totals of the tables this can also be done using reshape2. For example, taking the original table, we want to pivot it and sum each severity to get the total number of accidents per day.

```{r}
# Pivot table using reshape2 package
library(reshape2)
road_accidents_weekdays3 <- reshape2::dcast(road_accidents_small, Accident_Severity ~ Day_of_Week,
                                            value.var = "n", sum, margins = "Accident_Severity")
  
```

In this example, we use the margins argument to specify what we want to combine to create totals. So we want to add all the accident severity figures up for each weekday. Before using margin you need to specify how the margins are calculated, in this case we want a sum. Alternative options are to calculate the length, i.e. the number of rows.

```{r, echo = FALSE, results='asis'}

knitr::kable(head(road_accidents_weekdays3),
caption = "Number of road accidents by accident severity and weekday plus totals, reshape2::dcast")

```




The opposite can also be done using the melt function.

```{r}
# Pivot table using reshape2 package
library(reshape2)
road_accidents_melt <- reshape2::melt(road_accidents_weekdays2, id.vars = "Accident_Severity",
                                      measure.vars = c("1", "2", "3", "4", "5", "6", "7"),
                                      variable.name = "Day_of_Week", value.name = "n")
  
```

For the melt function you need to specify:


id.vars = "variables to be kept as columns"

measure.vars = c("variables to be created as one column")

variable.name = "name of created column using the measure.vars"

value.name = "name of value column"


```{r, echo = FALSE, results='asis'}
knitr::kable(head(road_accidents_melt),
caption = "Number of road accidents by accident severity and weekday, reshape2::melt")
```


## Dropping and selecting columns

Use the dplyr select function to both select and drop columns.

**Select columns**

```{r}
road_accidents_4_cols <- road_accidents %>%
  dplyr::select(acc_index, Accident_Severity, Date, Police_Force)
```

```{r, echo = FALSE, results='asis'}
knitr::kable(head(road_accidents_4_cols),
caption = "Four columns from road accidents 2017")
```


**Drop columns**

 Note that to drop columns the difference is putting a "-" in front of the variable name.
 
```{r}
road_accidents_3_cols <- road_accidents_4_cols %>%
  dplyr::select(-Police_Force)
```


```{r, echo = FALSE, results='asis'}
knitr::kable(head(road_accidents_3_cols),
caption = "Three columns from road accidents 2017")
```
 


## Filtering data

Use the dplyr filter function to filter data.

This example filters the data for slight severity accidents (accident severity 3).

```{r}
road_accidents_slight <- road_accidents_4_cols %>%
  dplyr::filter(Accident_Severity == 3)
```


```{r, echo = FALSE, results='asis'}
knitr::kable(head(road_accidents_slight),
caption = "Slight severity road accidents 2017")
```

To filter multiple conditions:

And operator
```{r}
road_accidents_filter <- road_accidents_4_cols %>%
  dplyr::filter(Accident_Severity == 3 & Police_Force == 4)
```

Or operator
```{r}
road_accidents_filter2 <- road_accidents_4_cols %>%
  dplyr::filter(Accident_Severity == 3 | Accident_Severity == 2)
```

**Note: filtering with characters must be wrapped in "quotations" e.g:**
```{r, eval = FALSE}
road_accidents_filter3 <- road_accidents %>%
dplyr::filter(`Local_Authority_(Highway)` == "E09000010")

```
Also note that in the above example the variable is quoted in back ticks (`). This is because some variable names confuse R due to brackets and numbers and need to be wrapped in back ticks so R knows that everything inside the back ticks is a variable name.


## Get counts of data

To get counts for groups of data, the dplyr tally function can be used in conjunction with the dplyr group by function. This groups the data into the required groups and then tallys how many records are in each group.

```{r}
# Create grouped data set with counts
road_accidents_small <- road_accidents %>%
  dplyr::group_by(Accident_Severity, Day_of_Week) %>%
  dplyr::tally()
```

The above example creates groups by accident severity and weekday and counts how many accidents are in each group (one row equals one accident therefore the tally is counting accidents).

```{r, echo = FALSE, results='asis'}
knitr::kable(head(road_accidents_small),
caption = "Road accidents 2017 by accident severity and weekday")
```



## Combine tables

When combining data from two tables there are two ways to do this in R:

* Bind the tables by basically either appending the tables on the rows or columns
* Join the tables using the dplyr version of SQL joins

**Binding tables**

Binding tables is mainly done to append tables by creating more rows, however tables can also be binded by adding more columns. Although it is recommended to use the dplyr join functions to combine columns (see 5.6).

```{r include=FALSE}
library(dplyr)

# create three tables for example
accidents_1 <- dplyr::filter(road_accidents_small, Accident_Severity == 1)
accidents_2 <- dplyr::filter(road_accidents_small, Accident_Severity == 2)
accidents_3 <- dplyr::filter(road_accidents_small, Accident_Severity == 3)

```

Here are three tables, one shows data for accident severity of 1, one for accident severity of 2, and one for accident severity of 3.

```{r, echo = FALSE, results='asis'}

knitr::kable(accidents_1, caption = "Number of fatal road accidents in 2017, by weekday")
knitr::kable(accidents_2, caption = "Number of serious injury road accidents in 2017, by weekday")
knitr::kable(accidents_3, caption = "Number of slight injury road accidents in 2017, by weekday")

```

To combine these tables we can use the bind_rows function from the dplyr package. Use bind_rows when you want to append the tables underneath one another to make one longer table, i.e. you want to add more rows.

**Ensure that the column names for each table are exactly the same in each table.**

```{r}
# combine tables using bind_rows
library(dplyr)

all_accidents <- accidents_1 %>%
  dplyr::bind_rows(accidents_2, accidents_3)
  
```


```{r, echo = FALSE, results='asis'}

knitr::kable(all_accidents, caption = "Road accident data 2017, bind_rows")

```

```{r include=FALSE}
library(dplyr)

# create two tables for example
road_acc_1 <- head(dplyr::select(road_accidents, acc_index, Police_Force, Accident_Severity))
road_acc_2 <- head(dplyr::select(road_accidents, acc_index, Date, Day_of_Week))

```


## Joining tables


Joins in R can be done using dplyr. This is generally to combine columns of data from two tables:


```{r}
# combine tables using left join
library(dplyr)

all_accidents_cols_join <- road_acc_1 %>%
  dplyr::left_join(road_acc_2, by = "acc_index")
```

This uses the same principles as SQL, by specifying what the tables should be joined on using the **by =** argument. 

```{r, echo = FALSE, results='asis'}

knitr::kable(all_accidents_cols_join, caption = "Road accident data 2017, left join")

```


Dplyr has all the usual SQL joins for example, inner_join, full_join, right_join. All of these are used in the same way as the left join example above.

Another useful join for data manipulation is an anti_join. This provides all the data that is not in the joined table. For example, the below snapshot of a table displays road accident totals broken down by accident severity and weekday:

```{r, echo = FALSE, results='asis'}

knitr::kable(head(road_accidents_small, caption = "Road accident data 2017 by accident severity and weekday"))

```

I am interested in creating two sub-groups of this data, a table for all accidents on a Monday (weekday 2), and all other accidents.

First, I get the Monday data using the dplyr filter function (see 5.3).

```{r include=FALSE}
library(dplyr)

# create filtered Monday table for example

accidents_monday <- dplyr::filter(road_accidents_small, Day_of_Week == 2)

```

```{r, echo = FALSE, results='asis'}

knitr::kable(head(accidents_monday, caption = "Road accident data 2017 on a Monday by accident severity"))

```

Then, I can use an anti-join to create a table which has all of the data that is not in the above table:

```{r}
# create table of all rows not in the joined table
library(dplyr)

all_accidents_not_monday <- road_accidents_small %>%
  dplyr::anti_join(accidents_monday, by = c("Accident_Severity", "Day_of_Week"))
```

The above code takes the initial table we want to get our data from (road_accidents_small) and anti joins accidents_monday. This says, "get all the rows from road_accidents_small that are not in accidents_monday". Again, note the need to specify what the join rows would be joined and compared by.


```{r, echo = FALSE, results='asis'}

knitr::kable(all_accidents_not_monday, caption = "Road accident data 2017 not on a Monday by accident severity")
```

## Select specific columns in a join

Doing a join with dplyr will join all columns from both tables, however sometimes not all columns from each table are needed.

Let's look at some previous tables again:

```{r, echo = FALSE, results='asis'}

knitr::kable(road_acc_1, caption = "Police force and accident severity information for accidents")
knitr::kable(road_acc_2, caption = "Date and weekday information for accidents")
```

Let's say we want acc_index and Police_Force from the first table, and Date from the second table.

```{r}
# select specific columns from each table and left join
library(dplyr)

road_acc_3 <- road_acc_1 %>%
  dplyr::select(acc_index, Police_Force) %>%
  dplyr::left_join(select(road_acc_2, acc_index, Date), by = "acc_index")
```

The above code takes the first table and uses the select statement to select the required columns from the first table. 

Then within the left_join command, to select the data from the second table, you again add the select statement.

**Note: you will need to select the joining variable in both tables but this will only appear once**

```{r, echo = FALSE, results='asis'}

knitr::kable(road_acc_3, caption = "Police force and Date information for specific accidents")
```

## Sum rows or columns

## Replace NAs or other values


## Creating new variables

## Summarising data

## Look up tables

