```{r include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

# Working with factors {#factors}

Working with factors can be tricky to both the new and the experience `R` user
as their behaviour is not always intuitive.  We can illustrate this with a 
couple of examples.  Firstly concatenating factors may not work as intended:

```{r}
my_factor <- factor(c("bob", "george", "bob"))
my_vector <- c(my_factor, my_factor)
my_vector
```

instead we would first need to convert to a character vector

```{r}
my_vector_2 <- c(as.character(my_factor), as.character(my_factor))
my_vector_2
```

Secondly, we may sometimes find a numeric vector is being stored as a factor:

```{r}
df <- read.csv("data/factor_example.csv")
my_vector <- df[[2]]
typeof(my_vector)
my_vector
```

to transform this to a numeric variable we can proceed as follows
```{r}
my_vector_2 <- as.numeric(levels(my_vector)[my_vector])
typeof(my_vector_2)
my_vector_2
```

To alleviate some of these issues it is useful to use the `forcats` package
**NOT FINISHED - need to write about `forcats`**
