```{r include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

# Working with factors {#factors}

## Common uses
Within the department there are three main ways you are likely to make use of 
factors:

* Tabulation of data (in particular when you want to illustrate zero occurences
of a particular value)
* Ordering of data for output (e.g. bars in a graph)
* Statistical models (e.g. in conjunction with `contrasts` when encoding 
categorical data in formulae)

### Tabulation of data
```{r}
# define a simple character vector
vehicles_observed <- c("car", "car", "bus", "car")
class(vehicles_observed)
table(vehicles_observed)

# convert to factor with possible levels
possible_vehicles <- c("car", "bus", "motorbike", "bicycle")
vehicles_observed <- factor(vehicles_observed, levels = possible_vehicles)
class(vehicles_observed)
table(vehicles_observed)
```

### Ordering of data for output
```{r}
# example 1
vehicles_observed <- c("car", "car", "bus", "car")
possible_vehicles <- c("car", "bus", "motorbike", "bicycle")
vehicles_observed <- factor(vehicles_observed, levels = possible_vehicles)
table(vehicles_observed)

possible_vehicles <- c("bicycle", "bus", "car", "motorbike")
vehicles_observed <- factor(vehicles_observed, levels = possible_vehicles)
table(vehicles_observed)

# example 2
df <- iris[sample(1:nrow(iris), 100), ]
ggplot(df, aes(Species)) + geom_bar()

df$Species <- factor(df$Species, levels = c("versicolor", "virginica", "setosa"))
ggplot(df, aes(Species)) + geom_bar()
```

### Statistical models
When building a regression model, R will automatically encode your independent 
character variables using `contr.treatment` contrasts.  This means that each 
level of the vector is contrasted with a baseline level (by default the first
level once the vector has been converted to a factor).  If you want to change 
the baseline level or use a different encoding methodology then you need to 
work with factors.  To illustrate this we use the `Titanic` dataset.

```{r}
# load data and convert to one observation per row
data("Titanic")
df <- as.data.frame(Titanic)
df <- df[rep(1:nrow(df), df[ ,5]), -5]
rownames(df) <- NULL
head(df)

# For this example we convert all variables to characters
df[] <- lapply(df, as.character)

# save to temporary folder
filename <- tempfile(fileext = ".csv")
write.csv(df, filename, row.names = FALSE)

# reload data with stringsAsFactors = FALSE
new_df <- read.csv(filename, stringsAsFactors = FALSE)
str(new_df)
```

First lets see what happens if we try and build a logistic regression model for
survivals but using our newly loaded dataframe

```{r, error = TRUE}
model_1 <- glm(Survived ~ ., family = binomial, data = new_df)
```

This errors due to the **Survived** variable being a character vector.  Let's 
convert it to a factor.

```{r}
new_df$Survived <- factor(new_df$Survived)
model_2 <- glm(Survived ~ ., family = binomial, data = new_df)
summary(model_2)
```

This works, but the baseline case for **Class** is `1st`.  What if we wanted it
to be `3rd`.  We would first need to convert the variable to a factor and choose
the appropriate level as a baseline

```{r}
new_df$Class <- factor(new_df$Class)
levels(new_df$Class)
contrasts(new_df$Class) <- contr.treatment(levels(new_df$Class), 3)
model_3 <- glm(Survived ~ ., family = binomial, data = new_df)
summary(model_3)
```

## Other things to know about factors
Working with factors can be tricky to both the new, and the experienced `R` 
user.  This is as their behaviour is not always intuitive.  Below we illustrate
three common areas of confusion

### Renaming factor levels
```{r, error=TRUE}
my_factor <- factor(c("Dog", "Cat", "Hippo", "Hippo", "Monkey", "Hippo"))
my_factor

# change Hippo to Giraffe
## DO NOT DO THIS
my_factor[my_factor == "Hippo"] <- "Giraffe"
my_factor

## reset factor
my_factor <- factor(c("Dog", "Cat", "Hippo", "Hippo", "Monkey", "Hippo"))

# change Hippo to Giraffe
## DO THIS
levels(my_factor)[levels(my_factor) == "Hippo"] <- "Giraffe"
my_factor
```

### Combining factors does not result in a factor
```{r}
names_1 <- factor(c("jon", "george", "bobby"))
names_2 <- factor(c("laura", "claire", "laura"))
c(names_1, names_2)

# if you want concatenation of factors to give a factor than the help page for
# c() suggest the following method is used:
c.factor <- function(..., recursive=TRUE) unlist(list(...), recursive=recursive)
c(names_1, names_2)

# if you only wanted the result to be a character vector then you could also use
c(as.character(names_1), as.character(names_2))
```

```{r, echo=FALSE}
# stop it messing up later code
rm(c.factor)
```

### Numeric vectors that have been read as factors
Sometimes we find a numeric vector is being stored as a factor
(a common occurence when reading a csv from Excel with #N/A values)

```{r}
# example data set
pseudo_excel_csv <- data.frame(names = c("jon", "laura", "ivy", "george"),
                               ages = c(20, 22, "#N/A", "#N/A"))
# save to temporary file
filename <- tempfile(fileext = ".csv")
write.csv(pseudo_excel_csv, filename, row.names = FALSE)

# reload data
df <- read.csv(filename)
str(df)
```

to transform this to a numeric variable we can proceed as follows
```{r check1234, error = TRUE}
df$ages <- as.numeric(levels(df$ages)[df$ages])
str(df)
```

## Helpful packages
If you find yourself having to manipulate factors often, then it may be worth
spending some time with the tidyverse package 
[forcats](https://forcats.tidyverse.org).  This was designed to make working
with factors simpler.  There are many tutorials available online but a good 
place to start is the official 
[vignette](https://forcats.tidyverse.org/articles/forcats.html).

